---
title: Reference-based analysis and clonal structure of tumor-specific human T cells
author: "M. Andreatta <massimo.andreatta at unil.ch> and S. Carmona <santiago.carmona at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
#output: html_notebook
---

# Introduction



# R environment

Install and load all required packages
```{r, message=FALSE, warning=F, results=FALSE}
renv::restore()

library(remotes)
library(BiocManager)

#BiocManager::install("scRepertoire")
#remotes::install_github("carmonalab/ProjecTILs")

library(ggplot2)
library(Seurat)
library(ProjecTILs)
library(scRepertoire)
library(GEOquery)
library(stringr)
library(patchwork)

# Increase timeout time for slow connections
options(timeout = max(900, getOption("timeout")))
```


# Data pre-processing

Download data from GEO
```{r}
geo_acc <- "GSE123813"
datadir <- "data"

cached_seurat <- sprintf("%s/Yost_alldata_seurat.rds", datadir)

#If we have already processed the data, read in a cached version
if(file.exists(cached_seurat)){
  query.object <- readRDS(cached_seurat)
} else {
  
  dir.create(datadir, showWarnings = FALSE)
  
  gse <- getGEO(geo_acc)
  getGEOSuppFiles(geo_acc, baseDir=datadir)
  
  #Read in expression matrix
  exp.mat <- read.delim(sprintf("%s/%s/GSE123813_bcc_scRNA_counts.txt.gz",
                                datadir, geo_acc), header=T, sep="\t")
  query.object <- CreateSeuratObject(counts = exp.mat, project = "Yost",
                                     min.cells = 3, min.features = 10)
  rm(exp.mat)
  
  #and the metadata associated with the GEX data
  meta <- read.delim(sprintf("%s/%s/GSE123813_bcc_all_metadata.txt.gz",
                             datadir, geo_acc), header = T, sep = "\t")
  meta.T <- read.delim(sprintf("%s/%s/GSE123813_bcc_tcell_metadata.txt.gz",
                               datadir, geo_acc), header = T, sep = "\t")
  
  query.object <- AddMetaData(query.object, meta$patient, col.name="patient")
  
  rownames(meta) <- meta$cell.id

  meta.T.c <- meta.T$cluster
  names(meta.T.c) <- meta.T$cell.id

  query.object <- AddMetaData(query.object, meta)
  query.object <- AddMetaData(query.object, meta.T.c, col.name = "cluster.T")
  
  #Read in TCR-seq data
  tcr.file <- sprintf("%s/%s/GSE123813_bcc_tcr.txt.gz", datadir, geo_acc)
  tcr.data <- read.csv(tcr.file, sep = "\t")
  
  #Check whether there is a alpha-beta pair
  alpha <- str_count(tcr.data$cdr3s_aa, pattern = "TRA:")
  beta <- str_count(tcr.data$cdr3s_aa, pattern = "TRB:")
  
  tcr.data$has_alphabeta <- ifelse(alpha>0 & beta>0, 1, 0)
  tcr.data$has_singlepair <- ifelse(alpha==1 & beta==1, 1, 0)
  
  query.object <- AddMetaData(query.object, tcr.data)
  
  #Make unique IDs for TCRs per patient
  query.object$cdr3s_pat <- NA
  inds <- !is.na(query.object$cdr3s_aa)
  query.object$cdr3s_pat[inds] <- paste0(query.object$cdr3s_aa[inds],
                                         "_",query.object$patient[inds])
  
  #Save object to disk
  saveRDS(query.object, cached_seurat)
}

```


```{r}
head(sort(table(query.object$cdr3s_pat), decreasing = T))
```


Subset on pre-treatment samples
```{r}
table(query.object$treatment, query.object$patient)

query.pre <- subset(query.object, subset=treatment=="pre")
```

# ProjecTILs analysis

## Load reference CD8 T map

Here we will project the query data onto a human reference TIL atlas for CD8 T cells, with the goal to interpret T cell diversity in the context of reference T cell subtypes. First, load the reference map:
```{r}
ref.file <- "CD8T_human_ref_v1.rds"
if (!file.exists(ref.file)) {
  download.file("https://figshare.com/ndownloader/files/38921366", destfile = ref.file)
}

ref.cd8 <- load.reference.map("CD8T_human_ref_v1.rds")
```

Let's have a look at the reference CD8 T cell subtypes
```{r}
DimPlot(ref.cd8, cols = ref.cd8@misc$atlas.palette, label = T) + theme(aspect.ratio = 1) +
    ggtitle("CD8 T reference") + NoLegend()
```


Show key marker genes for the reference subtypes

```{r fig.height=4, fig.width=3}
DefaultAssay(ref.cd8) <- "RNA"

genes <- c("SELL", "TCF7", "LEF1", "CCR7", "S1PR1", "LMNA", "IL7R", "GZMK", "FGFBP2",
    "FCGR3A", "XCL1", "XCL2", "CD200", "CRTAM", "GNG4", "TOX", "PDCD1", "HAVCR2",
    "GZMB", "PRF1", "LAG3", "KLRB1", "TRAV1-2")

VlnPlot(ref.cd8, features = genes, stack = TRUE, flip = TRUE, fill.by = "ident",
    cols = ref.cd8@misc$atlas.palette) + NoLegend()
ggsave("plots/CD8_ref_subset_violins.png", height=12, width=6)

DefaultAssay(ref.cd8) <- "integrated"
```

## Reference-based analysis

```{r}
#Split by patient
yost.list <- SplitObject(query.pre, split.by = "patient")

ncores = 6
yost.projected <- Run.ProjecTILs(yost.list, ref.cd8, ncores = ncores)
```



We can verify the expression profile using a panel of marker genes

```{r fig.height=4, fig.width=8}
genes4radar <- c("CD4", "CD8A", "TCF7", "CCR7", "IL7R", "LMNA", "GZMA", "GZMK", "FGFBP2",
    "XCL1", "CRTAM", "TOX", "PDCD1", "HAVCR2", "PRF1", "GZMB", "TRAV1-2",
    "KLRB1")

plot.states.radar(ref.cd8, query = yost.projected$su009, genes4radar = genes4radar)
```
Expression profile match very well those of the reference map



What is the subtype composition in different patients? Visualize projection and composition for individual patients:
```{r}
which.patient <- "su009"
a <- plot.projection(ref.cd8, query = yost.projected[[which.patient]],
                     pointsize = 0, linesize = 0.3) + NoLegend() +
                    ggtitle(which.patient)
b <- plot.statepred.composition(ref.cd8, query = yost.projected[[which.patient]]) 
a | b
```

Remove samples with very few cells. A minimum number of cells is required for robust analyses.
```{r fig.width=6}
sizes <- as.vector(lapply(yost.projected, ncol))

# At least 100 cells
keep <- names(sizes)[sizes > 100]

yost.projected <- yost.projected[keep]

plots <- lapply(names(yost.projected), function(x) {
  plot.statepred.composition(ref.cd8, query = yost.projected[[x]],
                             metric = "Percent") + ggtitle(x)
})
names(plots) <- names(yost.projected)

#sort by fraction of Tex cells
tex.fraction <- lapply(yost.projected, function(x) {
  t <- table(x$functional.cluster)
  unname(t["CD8.TEX"]/sum(t))
})
tex.fraction <- unlist(tex.fraction)

plots <- plots[names(sort(tex.fraction, decreasing = T))]

wrap_plots(plots, ncol=4)
```

Merge data into a single object to enable analyses across patients
```{r  fig.width=12, fig.height=8}
# First, merge data into single object
yost.merged <- Reduce(merge.Seurat.embeddings, yost.projected)
Idents(yost.merged) <- "functional.cluster"
```

# Clonotype analysis

See largest clonotypes in different patients.

First, calculate frequencies over the total of cells with a TCR chain in the same sample
```{r fig.width=5, fig.height=3}
freqs <- lapply(yost.projected, function(x) {
  table(x$cdr3s_pat) / sum(!is.na(x$cdr3s_pat))
})
freqs <- Reduce(c, freqs)
sorted <- sort(freqs, decreasing = TRUE)
largest.clones <- head(sorted, 6)
```

Plot individual clones
```{r fig.width=5, fig.height=3}
plots <- list()
for (i in 1:length(largest.clones)) {
  
  ctype <- names(largest.clones)[i]
  patient <- gsub("\\S+_","", ctype, perl=T)
  
  cells.in.clone <- which(yost.merged[["cdr3s_pat"]]==ctype)
  plots[[i]] <- plot.projection(ref.cd8, yost.merged[,cells.in.clone],
                                pointsize = 0.3, linesize = 0.3) + 
    ggtitle(sprintf("%s - Freq %.1f %%", patient, 100*largest.clones[i])) + NoLegend()
}

wrap_plots(plots, ncol = 3)

```

Summarize clonal sizes
```{r}
cutoffs <- c(0, 1, 5, 20, 50)
names(cutoffs) <- c("Single (=1)","Small (<=5)","Medium (<=20)","Large (<=50)","Hyper (>50)")

call <- "cdr3s_pat"
tab <- table(yost.merged[[call]])
yost.merged$clonesize <- NA
yost.merged$cloneType <- NA

for (r in 1:nrow(yost.merged[[]])) {
  e <- yost.merged@meta.data[r,call]
  if (!is.na(e) & e %in% names(tab)) {
    yost.merged@meta.data[r,"clonesize"] <- tab[e]
  }
}

for (i in seq_along(cutoffs)) {
  yost.merged$cloneType[yost.merged$clonesize > cutoffs[i]] <- names(cutoffs[i])
}
yost.merged$cloneType <- factor(yost.merged$cloneType, levels= rev(names(cutoffs)))

yost.merged$Frequency <- yost.merged$clonesize

#table(yost.merged$cloneType, yost.merged$clonesize)
```


See clonal composition and clonal diversity per cell subtype

```{r}
combined <- expression2List(yost.merged, split.by = "functional.cluster")

clonalDiversity(combined, cloneCall = "cdr3s_pat") +
  scale_color_manual(values = ref.cd8@misc$atlas.palette)

clonalProportion(combined, cloneCall = "cdr3s_pat", split = c(5, 10, 50, 100, 500, 2000))
  
clonalHomeostasis(combined, cloneCall = "cdr3s_pat",
                  cloneTypes = c(Rare=0.001, Small = 0.005, Medium = 0.01, Large = 0.1, Hyperexpanded= 1))

compareClonotypes(combined, 
                  clonotypes = names(largest.clones),
                  cloneCall="cdr3s_pat", 
                  graph = "alluvial")
```

```{r}
occupiedscRepertoire(yost.merged, x.axis = "functional.cluster")
```

Metrics of clonal overlap (e.g. Morisita index) can be used to assess clonal sharing between subtypes
```{r}
clonalOverlap(combined, cloneCall = "cdr3s_pat", method = "morisita") + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```

```{r}
library(ggraph)

clonalNetwork(yost.merged, 
              reduction = "umap", 
              identity = "functional.cluster",
              filter.clones = 1000,
              filter.identity = NULL,
              cloneCall = "cdr3s_pat") +
  scale_color_manual(values = ref.cd8@misc$atlas.palette)

clonalNetwork(yost.merged, 
              reduction = "umap", 
              identity = "functional.cluster",
              filter.clones = 1000,
              filter.identity = 'CD8.TEX',
              cloneCall = "cdr3s_pat") +
  scale_color_manual(values = ref.cd8@misc$atlas.palette)

```

Subset connectivity using circle graphs
```{r}
library(circlize)

circles <- getCirclize(yost.merged,
                       cloneCall = "cdr3s_pat",
                       group.by = "functional.cluster")

#Graphing the chord diagram
circlize::chordDiagram(circles,
                       self.link = 1, 
                       grid.col = ref.cd8@misc$atlas.palette)

```

Only on large clones
```{r}
large <- subset(yost.merged, subset = clonesize >= 5)

circles <- getCirclize(large,
                       cloneCall = "cdr3s_pat",
                       group.by = "functional.cluster")

#Graphing the chord diagram
circlize::chordDiagram(circles,
                       self.link = 1, 
                       grid.col = ref.cd8@misc$atlas.palette)

```

Clonotype bias - find clones with significant enrichment for specific cell subtypes

```{r}
t <- table(yost.merged$patient)
large.patients <- names(t)[t > 300]

yost.large <- subset(yost.merged, subset = patient %in% large.patients)

clonotypeBias(yost.large, cloneCall = "cdr3s_pat", split.by = "patient",
              group.by = "functional.cluster", min.expand = 5, n.boots = 20) +
  scale_color_manual(values = ref.cd8@misc$atlas.palette)

biased <- clonotypeBias(yost.large, cloneCall = "cdr3s_pat", split.by = "patient",
              group.by = "functional.cluster", min.expand = 5, exportTable = TRUE)

most.biased <- biased[order(biased$Z.score, decreasing = TRUE),]
head(most.biased)
```

```{r fig.width=5, fig.height=3}
plots <- list()
for (i in 1:6) {
  
  ctype <- most.biased[i, "Clone"]
  patient <- most.biased[i, "Sample"]
  size <- most.biased[i, "ncells"]
  zscore <- most.biased[i, "Z.score"]
  cells.in.clone <- which(yost.merged[["cdr3s_pat"]]==ctype)
  title <- sprintf("Clone %s, size %s, Z=%.1f", i, size, zscore)
  
  plots[[i]] <- plot.projection(ref.cd8, yost.merged[,cells.in.clone],
                                pointsize = 0.3, linesize = 0.3) + 
    ggtitle(title) + NoLegend()
}

wrap_plots(plots, ncol = 3)

```
